---
title: "ChromBPNet Model Statistics - HSC PCW"
author: "Mervin M Fansler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document: 
    code_folding: show
    df_print: paged
    theme: cosmo
    toc: true
    toc_float: true
---

# Purpose

This analysis summarizes the fit statistics for the ChromBPNet model trainings of HSC models in developmental time.

Text under each metric is adapted from the ChromBPNet training and quality check report.

# Initialization

## Libraries
```{r libs, message=FALSE, warning=FALSE}
library(jsonlite)
library(magrittr)
library(tidyverse)
library(showtext)
```

## Parameters
```{r set_params}
set.seed(20251103)

FILES_JSON_PATTERN="/work/aaa/projects/chrombpnet-devmult/pipeline/results/chrombpnet_nobias/pretrained_bias/HSC_PCW*/fold_*/evaluation/*._chrombpnet_metrics.json"

CSV_FRAGS="/work/aaa/projects/chrombpnet-devmult/scratch/frags/hsc_pcw_fragcts.csv"

COLS_PCW=c(
  "HSC_PCW5"="#FFBF00",
  "HSC_PCW6"="#ff986d",
  "HSC_PCW7"="#ff8782",
  "HSC_PCW8"="#ff7b9b",
  "HSC_PCW9"="#f176b4",
  "HSC_PCW10"="#d97fce",
  "HSC_PCW11"="#b989e2",
  "HSC_PCW12"="#9094ed",
  "HSC_PCW13"="#60a4f2",
  "HSC_PCW14"="#32b1eb",
  "HSC_PCW15"="#29badb",
  "HSC_PCW16"="#4bc0c8",
  "HSC_PCW17"="#8dd7db",
  "HSC_PCW18"="#c0dedf")

theme_set(theme_bw())
```


# Data
## Loading
```{r load_data, message=FALSE}
df_frags <- read_csv(CSV_FRAGS) %>%
  mutate(sample_id=fct(sample_id, levels=names(COLS_PCW)))

df_metrics <- tibble(file=Sys.glob(FILES_JSON_PATTERN)) %>%
  mutate(sample=str_match(file, "([^/]+)/fold")[,2]) %>%
  filter(sample %in% names(COLS_PCW)) %>%
  mutate(sample=fct(sample, levels=names(COLS_PCW))) %>%
  mutate(fold=str_match(file, "fold_([0-9]+)/")[,2]) %>%
  mutate(raw_metrics=map(file, fromJSON)) %>%
  mutate(spearmanr=sapply(raw_metrics, function (l) l$counts_metrics$peaks$spearmanr),
         pearsonr=sapply(raw_metrics, function (l) l$counts_metrics$peaks$pearsonr),
         mse=sapply(raw_metrics, function (l) l$counts_metrics$peaks$mse),
         median_jsd=sapply(raw_metrics, function (l) l$profile_metrics$peaks$median_jsd),
         median_norm_jsd=sapply(raw_metrics, function (l) l$profile_metrics$peaks$median_norm_jsd)) %>%
  mutate(raw_metrics=NULL)
```

# Analysis

## Total Fragments in Peaks
This is a summary of the number of fragments present in the peaks used for model training. Note this is overall and not per fold.

```{r nfrags, fig.width=2.96, fig.height=2.0}
df_frags %>%
  ggplot(aes(x=sample_id, y=n_frags/1e6, fill=sample_id)) +
  geom_bar(stat='identity', color='black', linewidth=0.3) +
  scale_fill_manual(values=COLS_PCW, guide='none') +
  scale_x_discrete(guide=guide_axis(angle=90), labels=function(x) sub("^HSC_PCW", "PCW ", x)) +
  scale_y_continuous(expand=c(0,0,0.1,0), trans="log10") +
  labs(x=NULL, y="Fragments in Peaks (M)") +
  theme(axis.title.y=element_text(size=8))

ggsave("figures/suppl-fig-chrombpnet-nfragments-hsc-pcw.pdf", dpi=300)
```

## Counts Metrics
Counts metrics characterize how well the predicted counts in peaks reproduce the observed counts.

### Spearman's rho - Overall
Higher Spearman's rho in peaks is better.

```{r spearmanr, fig.width=3, fig.height=2.0}
df_metrics %>%
  ggplot(aes(x=sample, y=spearmanr)) +
  stat_summary(aes(fill=sample), fun='mean', geom='bar', color='black', linewidth=0.3) +
  stat_summary(fun.data=mean_se, fun.args=list(mult=2), geom='errorbar', color='black', linewidth=0.5, width=0.7) +
  geom_jitter(height=0, width=0.15, size=0.5) +
  scale_fill_manual(values=COLS_PCW, guide='none') +
  scale_x_discrete(guide=guide_axis(angle=90), labels=function(x) sub("^HSC_PCW", "PCW ", x)) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  labs(x=NULL, y="Spearman's ρ") +
  theme(axis.title.y=element_text(size=8))

showtext_auto()
ggsave("figures/suppl-fig-chrombpnet-metric-spearman-hsc-pcw.pdf", dpi=300)
showtext_auto(FALSE)
```


### Pearson's rho - Overall
The Pearson's rho in peaks should be greater than 0.5 (higher the better).

```{r pearsonr, fig.width=3, fig.height=2.0}
df_metrics %>%
  ggplot(aes(x=sample, y=pearsonr)) +
  stat_summary(aes(fill=sample), fun='mean', geom='bar', color='black', linewidth=0.3) +
  stat_summary(fun.data=mean_se, fun.args=list(mult=2), geom='errorbar', color='black', linewidth=0.5, width=0.7) +
  geom_jitter(height=0, width=0.15, size=0.5) +
  scale_fill_manual(values=COLS_PCW, guide='none') +
  scale_x_discrete(guide=guide_axis(angle=90), labels=function(x) sub("^HSC_PCW", "PCW ", x)) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  labs(x=NULL, y="Pearson's ρ") +
  theme(axis.title.y=element_text(size=8))

showtext_auto()
ggsave("figures/suppl-fig-chrombpnet-metric-pearson-hsc-pcw.pdf", dpi=300)
showtext_auto(FALSE)
```

### MSE - Overall
MSE (Mean Squared Error) should be low in peaks. 

```{r mse, fig.width=3, fig.height=2.0}
df_metrics %>%
  ggplot(aes(x=sample, y=mse)) +
  stat_summary(aes(fill=sample), fun='mean', geom='bar', color='black', linewidth=0.3) +
  stat_summary(fun.data=mean_se, fun.args=list(mult=2), geom='errorbar', color='black', linewidth=0.5, width=0.7) +
  geom_jitter(height=0, width=0.15, size=0.5) +
  scale_fill_manual(values=COLS_PCW, guide='none') +
  scale_x_discrete(guide=guide_axis(angle=90), labels=function(x) sub("^HSC_PCW", "PCW ", x)) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  labs(x=NULL, y="MSE") +
  theme(axis.title.y=element_text(size=8))

ggsave("figures/suppl-fig-chrombpnet-metric-mse-hsc-pcw.pdf", dpi=300)
```

## Profile Metrics
Profile metrics are about how well the predicted shapes of peaks reproduce the observed shapes.

### JSD - Overall
Lower median JSD (Jensen Shannon Divergence between observed and predicted) is better. JSD is sensitive to read-depth.
Higher read-depth results in better metrics.

```{r jsd, fig.width=3.17, fig.height=2.0}
df_metrics %>%
  ggplot(aes(x=sample, y=median_jsd)) +
  stat_summary(aes(fill=sample), fun='mean', geom='bar', color='black', linewidth=0.3) +
  stat_summary(fun.data=mean_se, fun.args=list(mult=2), geom='errorbar', color='black', linewidth=0.5, width=0.7) +
  geom_jitter(height=0, width=0.15, size=0.5) +
  scale_fill_manual(values=COLS_PCW, guide='none') +
  scale_x_discrete(guide=guide_axis(angle=90), labels=function(x) sub("^HSC_PCW", "PCW ", x)) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  labs(x=NULL, y="Jensen-Shannon\nDivergence (median)") +
  theme(axis.title.y=element_text(size=8))

ggsave("figures/suppl-fig-chrombpnet-metric-jsd-hsc-pcw.pdf", dpi=300)
```


### Normalized Median JSD - Overall
Median norm JSD is median of the min-max normalized JSD where min JSD is the worst case JSD, i.e, JSD of observed with 
uniform profile and max JSD is the best case JSD, i.e, 0. Median norm JSD is higher the better. Median norm JSD is 
sensitive to read-depth. Higher read-depth results in better metrics.

```{r norm_jsd, fig.width=3.34, fig.height=2.0}
df_metrics %>%
  ggplot(aes(x=sample, y=median_norm_jsd)) +
  stat_summary(aes(fill=sample), fun='mean', geom='bar', color='black', linewidth=0.3) +
  stat_summary(fun.data=mean_se, fun.args=list(mult=2), geom='errorbar', color='black', linewidth=0.5, width=0.7) +
  geom_jitter(height=0, width=0.15, size=0.5) +
  scale_fill_manual(values=COLS_PCW, guide='none') +
  scale_x_discrete(guide=guide_axis(angle=90), labels=function(x) sub("^HSC_PCW", "PCW ", x)) +
  scale_y_continuous(expand=c(0,0,0.1,0)) +
  labs(x=NULL, y="Jensen-Shannon\nDivergence\n(median normalized)") +
  theme(axis.title.y=element_text(size=8))

ggsave("figures/suppl-fig-chrombpnet-metric-normjsd-hsc-pcw.pdf", dpi=300)
```

# Conclusion

All models are very well-fitted for counts. Some of the PCWs had higher JSD for profile metric 
(e.g., PCW 5, 12, 15), but that is consistent with having lower coverage for these.

---

# Runtime Details
## Session Info
<details>
```{r sesh_info, echo=FALSE}
sessionInfo()
```
</details>

## Conda Environment
<details>
```{bash conda_info, comment="", echo=FALSE}
if ! command -v conda &> /dev/null
then
  echo "Conda not detected."
elif [ -z "${CONDA_PREFIX}" ]
then
  echo "No active Conda environment."
else
  echo "## Conda Environment YAML"
  PREFIX=$(dirname $(dirname $(which R)))
  conda env export -p $PREFIX
fi
```
</details>
