---
title: "11.compsotion_Poisson"
---

```{r knitr, include = FALSE}
DOCNAME <- "11.compsotion_Poisson"
dir.create(here::here("output", DOCNAME), showWarnings = FALSE)

NOW <- Sys.time()
# Time chunks during knitting
knitr::knit_hooks$set(timeit = function(before) {
  if (before) {
    print(paste("Start:", Sys.time()))
    NOW <<- Sys.time()
  } else {
    print(paste("Stop:", Sys.time()))
    print(Sys.time() - NOW)
  }
})

knitr::opts_chunk$set(
  autodep        = TRUE,
  cache          = FALSE,
  cache.lazy     = FALSE,
  cache.comments = FALSE,
  echo           = TRUE,
  error          = FALSE,
  fig.align      = "center",
  fig.width      = 10,
  fig.height     = 8,
  message        = FALSE,
  warning        = FALSE,
  timeit         = TRUE
)
```

```{r}
library(tidyverse)
library(data.table)

# plot
library(scales)
library(scattermore)
library(ggrastr)
library(ggpubr)
library(ggbeeswarm)

# patch
library(patchwork)
library(cowplot)
mytheme <- theme_cowplot(
  font_size = 10,
  rel_small = 8 / 10,
  rel_tiny = 8 / 10,
  rel_large = 10 / 10
)

# color
library(ggsci)

# srat
library(Seurat)

# model
library(lme4)
library(numDeriv)
library(DHARMa)
library(lmtest)
```

```{r source, cache = FALSE}
#source(here::here("utils/preprocess.R"))
source(here::here("utils/markers.R"))
source(here::here("utils/plot.R"))
theme_set(publ_theme)
source("/work/DevM_analysis/utils/colors.R")
source(here::here("utils/CellTypeCompositionAnalysis.R"))
```

## Set up

Load cellmeta

```{r}
srat <- readRDS("/work/DevM_analysis/02.abundance/Milo_FL_HSC_PCW250401/data/FL_rna_clustered.rds")
srat
```

load HSC sub-clustering

```{r}
df_hsc = read.csv('/work/DevM_analysis/01.annotation/11.subclustering/HSC/data/FL_wnn_cellmeta.v00.csv') |> 
  dplyr::select(barcode = X, anno_wnn_hsc = leiden_wnn_0.3) |> 
  mutate(anno_wnn_hsc = paste0("HSC-", anno_wnn_hsc)) |> 
  mutate(anno_wnn_hsc = factor(anno_wnn_hsc, levels = paste0("HSC-", 0:4)))
head(df_hsc)
```

Neat

```{r}
obs_tbl <- srat@meta.data |>
  rownames_to_column("barcode") |> 
  left_join(df_hsc, by = "barcode")

obs_tbl$PCWsca <- scale(as.integer(as.character(obs_tbl$PCW)))
unique(obs_tbl$PCWsca)
```

```{r}
metadata <- obs_tbl |> 
  dplyr::select(sampleID, libraryID, donorID, Sex, PCW, PCWsca, Batch) |> 
  distinct() |> 
  mutate(Sex = factor(Sex, levels = c("M", "F")),
         Batch = factor(Batch),
         donorID =factor(donorID))
str(metadata)
```

count table (cell number of each cell type and each sample)

```{r}
Y <- .make_count_matrix(obs_tbl, "sampleID", "anno_wnn_hsc")
dim(Y)
```

input for glmer

```{r}
input_dbl <- .make_input_for_glmer(Y = Y, metadata_tbl = metadata, colSample = "sampleID")
head(input_dbl)
```

## Library/Donor/Batch as random, Sex as fixed (model 4)

`Ncs ∼ PCW+Sex+(1∣Celltype)+(1∣Library)+(1∣Donor)+(1∣Batch)+(PCW−1∣Celltype)+(1∣Library:Celltype)+(1∣Donor:Celltype)+(1∣Batch)`

```{r}
terms <- c(
    "PCWsca",
    "Sex",
    sprintf("(1|%s)", c("libraryID", "donorID", "Batch")),
    sprintf("(%s-1|Celltype)", "PCWsca"),
    sprintf("(1|%s:Celltype)", c("libraryID", "donorID", "Batch"))
  )
formula_str <- paste(c("I(Y) ~ (1|Celltype)", terms), collapse = "+")
f <- as.formula(formula_str)
f
```

fit

```{r eval=FALSE}
res4 <- CellTypeCompositionAnalysis2(f = f, data = input_dbl)
saveRDS(res4, file = here::here("output", DOCNAME, "GLMM.Library_Donor_Batch_asRandom.Sex_asFixed.rds"))
```

```{r}
res4 <- readRDS(here::here("output", DOCNAME, "GLMM.Library_Donor_Batch_asRandom.Sex_asFixed.rds"))
summary(res4$res.prop)
```

```{r fig.width=4, fig.height=6}
plt_gen <- plot_ranef(res4$ranef, vars=list(PCWsca = c('Slope-PCWsca')), 
                     celltypes = levels(obs_tbl$anno_wnn_hsc),
                     celltype_order = rev(levels(obs_tbl$anno_wnn_hsc)),
                     highlightLtsr = 0.95) + theme_pubr(legend = "right") +
   theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(),
         axis.text.x=element_text(angle=60, vjust=1, hjust=1))
plt_gen

ggsave2(
  filename = here::here("output", DOCNAME, "Dotplot.Library_Donor_Batch_asRandom.Sex_asFixed.pdf"),
  width = 4, height = 12
)
```

```{r fig.width=10, fig.height=6}
plt_gen <- plot_ranef(res4$ranef, vars=list(PCWsca = c('Slope-PCWsca'),
                                            Batch = levels(metadata$Batch)), 
                     celltypes = levels(obs_tbl$anno_wnn_hsc),
                     celltype_order = rev(levels(obs_tbl$anno_wnn_hsc)),
                     highlightLtsr = 0.95) + theme_pubr(legend = "right") +
   theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(),
         axis.text.x=element_text(angle=60, vjust=1, hjust=1))
plt_gen

ggsave2(
  filename = here::here("output", DOCNAME, "Dotplot.Library_Donor_Batch_asRandom.Sex_asFixed.FULL.pdf"),
  width = 10, height = 12
)
```

Plot variable explained variance

```{r fig.width=6, fig.height=4}
plot_sdse(sdse_tbl = res4$sdse, colSample = "sampleID") +
  mytheme
```

### residual diagnostics

```{r fig.width=8, fig.height=5}
sim_res <- simulateResiduals(fittedModel = res4$res.prop)
plot(sim_res)
```

dispersion test

```{r fig.width=5, fig.height=4}
testDispersion(sim_res)
```

Test for Zero-Inflation

```{r fig.width=5, fig.height=4}
testZeroInflation(sim_res)
```

## Plot together with milo

WNN-milo

```{r}
df_milo <- read.csv("/work/DevM_analysis/02.abundance/Milo_FL_HSC_PCW250401/data/FL_wnn_milo.nhood_test_res.csv")
dim(df_milo)
table(df_milo$nhood_annotation)
table(df_milo$donorSpecific)
```

Milo plot

```{r fig.width=5, fig.height=4}
d4p <- df_milo |> 
  filter(nhood_annotation != "Mixed", donorSpecific == "Good") |> 
  group_by(nhood_annotation) %>%
  mutate(m_logFC = median(logFC)) |> 
  ungroup() |> 
  arrange(-m_logFC) |> 
  mutate(nhood_annotation=factor(nhood_annotation, levels=unique(nhood_annotation))) |> 
  mutate(logFC_color = ifelse(is_signif=="True", logFC, NA))

p_milo <- d4p %>%
  ggplot(aes(x = nhood_annotation, y = logFC, color=logFC_color)) +
  geom_quasirandom(size=0.8, width = .3) +
  geom_point(data= . %>% distinct(nhood_annotation, m_logFC), aes(x=nhood_annotation, y=m_logFC), color="black")  +
  scale_color_gradient2() +
  coord_flip() +
  guides(color="none", fill="none") +
  geom_hline(yintercept=0, linetype=2) +
  xlab("") + ylab("Log Fold Change in time") +
  theme_pubr(legend = "right", base_size = 15)
p_milo
```

GLMM

```{r fig.width=4, fig.height=5}
plt_gen <- plot_ranef2(res4$ranef, vars=list(PCWsca = c('Slope-PCWsca')), 
                     celltypes = intersect(obs_tbl$anno_wnn_hsc, levels(d4p$nhood_annotation)),
                     celltype_order = levels(d4p$nhood_annotation),
                     highlightLtsr = 0.95) +
  labs(y = NULL, x = "PCW") +
  theme_pubr(legend = "right", base_size = 15) +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank(),
        axis.text.x=element_blank())
plt_gen
```

Align

```{r fig.width=7, fig.height=4}
p_milo +
  (plt_gen + theme(axis.text.y = element_blank())) +
  plot_layout(widths = c(3, 1))

ggsave2(
  filename = here::here("output", DOCNAME, "HSC.Ctype_DA_milo_and_GLMM.pdf"),
  width = 7, height = 4
)
```

## Session info

