---
title: "HSC.dev_de_lmm_qScoreAdj.rmd"
output:
  html_document:
    css: style.css
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
    code_folding: show
    df_print: paged
    number_sections: false
    self_contained: true
    highlight: kate
---

```{r knitr,include = FALSE}
# Time chunks during knitting
knitr::knit_hooks$set(timeit = function(before) {
  if (before) {
    print(paste("Start:", Sys.time()))
    NOW <<- Sys.time()
  } else {
    print(paste("Stop:", Sys.time()))
    print(Sys.time() - NOW)
  }
})
knitr::opts_chunk$set(
  autodep        = TRUE,
  cache          = FALSE,
  cache.lazy     = FALSE,
  cache.comments = FALSE,
  echo           = TRUE,
  error          = FALSE,
  fig.align      = "center",
  fig.width      = 10,
  fig.height     = 8,
  message        = FALSE,
  warning        = FALSE,
  timeit         = TRUE
)
```

>* LMM of gene expression along developmental time (pcw), with quiescence module score included as a covariate to adjust for quiescent cell-state effects. And compare with previous LMM model without adjusting for quiescence.
>* Model: exp ~ logUMI + logFeature + percent.mt + Batch + libraryID +
                 donorID + Sex + module_score(BM_LTvsST_up) + scaled_pcw
>* Thresholds for significant genes: LTSR > 0.9


```{r, message=FALSE}
library(ggplot2)
library(tidyverse)
library(ComplexUpset)
library(UpSetR)
library(eulerr)
library(ggsci)
library(ggrepel)
library(patchwork)

workdir <- '/work/Jupyterlab/Project/CellCycle/03.Disentanglement/'
datadir <- paste0(workdir,'data/')
plotdir <- paste0(workdir,'plot/')
```



## Load data

```{r load data, message=FALSE}
lmm <- readRDS('/work/DevM_analysis/data/HSC_dev_diff_genes/HSC_lmm_de_time_df.rds')
lmm_adjQ <- readRDS(paste0(datadir,'111.HSC_adjQ_lmm_de_time_df.rds'))
lmm_adjQ <- lmm_adjQ[rownames(lmm),]
stopifnot(rownames(lmm)==rownames(lmm_adjQ))
```


## Beta Correlation

Pearson correlation on beta values of all genes

```{r, message=FALSE}
cor.test(lmm$beta,lmm_adjQ$beta,method = "pearson")
```

```{r,message=FALSE,fig.width=5,fig.height=5,fig.align='center',class.source = "fold-hide"}
d4p <- lmm |> left_join(lmm_adjQ,by='gene')
ggplot(d4p,aes(x=beta.x,y=beta.y)) +
  geom_point(size=0.5) +
  geom_abline(colour = 'darkred',linetype = 'dashed',slope = 1, intercept = 0) +
  scale_x_continuous(breaks = seq(-0.4,0.4,0.2)) +
  scale_y_continuous(breaks = seq(-0.4,0.4,0.2)) +
  labs(x='Beta value\n(lmm without quiescence adjusted)',y='Beta value\nlmm with quiescence adjusted') +
  theme(panel.background = element_blank(),
        panel.border = element_rect(color='black',fill = NA),
        panel.grid = element_line(linetype = 'dotted',color = 'lightgrey')) +
  annotate("text",
           x = max(d4p$beta.x, na.rm = TRUE),
           y = min(d4p$beta.y, na.rm = TRUE),
           label = "ρ = 0.96, p < 2.2e-16",
           hjust = 1, vjust = 0)
```


## Overlapping {.tabset}

Overlapping genes identified as significant across development in both models

```{r, message=FALSE}
lmm_sig <- lmm |> filter(ltsr>0.9)
lmm_adjQ_sig <- lmm_adjQ |> filter(ltsr>0.9)
listInput<- list(
  'lmm_up'=lmm_sig[lmm_sig$beta>0,'gene'],
  'lmm_dn'=lmm_sig[lmm_sig$beta<0,'gene'],
  'lmm_adjQ_up'=lmm_adjQ_sig[lmm_adjQ_sig$beta>0,'gene'],
  'lmm_adjQ_dn'=lmm_adjQ_sig[lmm_adjQ_sig$beta<0,'gene']
)

```

Among genes identified in the unadjusted model:<br>
Nearly **70%** genes remain significant after adjusting for quiescence, indicating developmental-driven genes.<br>
While nearly **30%** genes lose significance, suggesting confounding by quiescence population.
<br>

### Venn

```{r, message=FALSE,fig.width=20,fig.height=8,fig.align='center',class.source = "fold-hide"}
fit <- euler(listInput)
plot(
  fit,
  fills = c("#4688ECFF","#FCA700FF","#94C748FF","#C97CF4FF"),
  alpha = 0.7,
  reflect = TRUE,
  quantities = list(cex = 1.2),
  labels = list(cex = 1.5),
  adjust_labels = TRUE,
  reflect = TRUE
)
```

### Upset

```{r, message=FALSE,fig.width=7,fig.height=7,fig.align='center',class.source = "fold-hide"}
ComplexUpset::upset(fromList(listInput),
      sort_sets=FALSE,
      intersect = names(listInput),
      width_ratio = 0.2)
```

```{r,include=FALSE,eval=FALSE}
p <- ComplexUpset::upset(fromList(listInput),
      sort_sets=FALSE,
      intersect = names(listInput),
      width_ratio = 0.2)
Cairo::CairoPDF(paste0(plotdir,'111.UpsetR.pdf'),
                width = 10,height = 10,family = 'Arial')
p
dev.off()
```

